Think deep for reef resilience

### 2025-09-15 paper/paper_draft.qmd
Regenerate figures with outliers removed
Update reef counts throughout paper and reduce emphasis on the connectivity - depth interactions in results/discussion

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress
    events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour. Our findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest as
the predictors, including the median reef depth, mean DHW for the given GCM, log10-weighted
incoming connectivity, log10-weighted outgoing connectivity, and log10-outgoing-to-incoming
connectivity ratio. The full dataset consisted of the time series cluster assignment of each
of the 2197 reefs under the five GCMs explored, for a total of 12,145 samples. The dataset **[]
was randomly split into training (60%) and testing (40%) subsets to ensure model
generalisability. The RF classifier achieved 79% accuracy on the training set and 59% on the
test set.

A 3-fold cross-validation was also conducted, arriving at a similar 57% accuracy rate.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.68, 0.43 and 0.62 respectively, with precision of 0.77, 0.49 and 0.66 when evaluated with
the test set (consisting of 4858 samples). Lower performance with the Medium cluster
reflects the inherent difficulty of classifying transitional states in natural systems. The
consistent level of accuracy across test and cross-validation demonstrates the generalised
performance of the RF classifier, while not accurate, was deemed sufficiently informative
for the purpose of assessing the influence of reef characteristics.

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its cluster assignment
into Low, Medium, or High. Bioregion classification was excluded from this analysis as a
predictor due to its categorical nature providing limited interpretable information, and is
further justified by its low feature importance score of 9%, the lowest of all predictors
(@fig-ts-cluster-pdp; Panel A). Reef median depth, mean DHW level and weighted incoming
connectivity (Log10) have the highest feature importance scores, accounting for 22%, 17% and
14% predictive importance respectively. The carrying capacity of a reef accounts for 12%
feature importance, while outgoing connection strength, source-to-sink ratio and larval
retention account for 10%, 9% and 7% respectively.

@@

Analyses for source-to-sink ratio and incoming and outgoing connectivity metrics suggest
that reefs with greater outgoing connectivity and lower incoming connectivity levels have
higher probability of being clustered into the Low cluster category compared to other reefs.
Additionally, two-way analyses revealed an interaction between reef depth and incoming
connectivity in determining cluster category assignment, particularly at the high depth
range (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses revealed that reefs at greater
than 10m depth are unlikely to be clustered into the Low cluster category, regardless of
their incoming connectivity levels. However, at greater than 10m depth reefs with higher
incoming connectivity are more likely to be assigned to a High cluster, while reefs with low
incoming connectivity are more likely to be assigned to a Medium cover cluster
(@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses, for depth and outgoing connection
strength, revealed a slight interaction with deep reefs with high outgoing connectivity
having the greatest probability of being clustered into a high cover cluster
(@sfig-ts-cluster-pdp-twoway-outgoing).

@@

Current reef restoration and intervention efforts on the GBR, such as deployment of coral
seeding, coral fragment devices and Crown of Thorns Starfish control, target reefs with
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, intervention depth is likely constrained by
snorkelling and diving deployment and monitoring logistical accessibility, visual
identification of sites from the surface, in combination with tourism return value. Our
results suggest that of the assessed reefs occurring within the 1-10m depth range (1551
reefs), 59% (920) of reefs are consistently clustered into the Low and Medium cover clusters
across the GCMs (at the GBRMPA management region scale), with many unable to maintain
positive carbonate budgets under future conditions. Partial dependence analysis revealed a
range of depths from 10-15m where the probability of a reef being clustered into the High
cover cluster increased substantially. Additionally, two-way partial dependence analysis
revealed that high incoming connectivity is important for High cluster assignment for reefs
at depths of \>10m. Considering this 10-15m depth range when assessing possible intervention
sites may reduce the exposure of planted and seeded corals to high thermal stress, while
still maintaining effective logistical access to sites. Managers are also likely to consider
the potential for positive flow on effects due to outgoing connectivity when selecting
intervention sites [@hock2017]. Two-way partial dependence analysis revealed reefs with high
depths and high outgoing connectivity levels have a high probability of being clustered into
more resilient reef clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results suggest
that interventions targetting deep reefs will also benefit from high incoming and outgoing
connectivity levels, allowing for high potential for site resilience and outgoing flow of
benefits and recovery. Transient, and difficult to measure, properties such as reef
connectivity are likely to increase uncertainty in intervention planning to a greater degree
than environmental characteristics such as reef depth. Overall, reef intervention planning
is a complex and challenging decision-making process [@anthony2020], and our results suggest
that intervention site depth may be an important factor to additionally consider when
assessing the long-term resilience of sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Our results suggest that incoming connectivity contributes positively to high
cover cluster allocation for reefs over 10m deep. At depths of less than 10m thermal stress
may be too high for incoming juveniles to contribute to long term recovery. While there is
interaction between incoming connectivity and depth for reef cluster prediction, according
to available input data, there are few reefs in the GBR that occur at depths \>10m and have
Log10 weighted incoming connectivity levels below 0 (@sfig-access-esm1-5-mean-carb-budget),
resulting in overall low importance of connectivity as an individual feature for clustering
across the GBR. Furthermore, our results suggest no direct, first order, relationship
between weighted incoming connectivity and the ability of reefs to maintain positive
carbonate budgets under future conditions. There has been contrasting evidence [@bode2018;
@hock2017] surrounding whether recent generation connectivity modelling is able to identify
reefs that are important for recovery processes, with findings dependent on the larval
connectivity model used. These investigations were performed using prior generations of
larval connectivity models parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018]
time periods, without considering the influence of reef depth on resilience. Using updated
1km^2^ resolution larval connectivity across the whole GBR, with a limited temporal sample
of particle flows, our results suggest that larval connectivity alone may not be important
or reef resilience and recovery, but should be considered in combination with more
influential factors such as depth. Comprehensive comparisons and validation of multiple
connectivity model outputs would improve certainty and allow for the targetting of reefs by
combining abiotic factors such as depth with connectivity.

@@

The results presented suggest that when modelled as part of a large and complex ecosystem,
reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in
support of depth as a form of thermal refugia for shallow-water reefs on the GBR. Targetting
reefs with median depths greater than 10m and high incoming and ougoing larval connections
may aid in identifying the most resilient reefs under climate change. We suggest further
analyses that investigate the benefits of incorporating depth-based resilience into reef
intervention decision making, as well as further research and development in connectivity
modelling methods for whole-ecosystem scales.

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress
    events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour at 15m depth. Our findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest as
the predictors, including the median reef depth, mean DHW for the given GCM, log10-weighted
incoming connectivity, log10-weighted outgoing connectivity, and log10-outgoing-to-incoming
connectivity ratio. The full dataset consisted of the time series cluster assignment of each
of the 2197 reefs under the five GCMs explored, for a total of 10,985 samples. The dataset
\*\*\[\] was randomly split into training (60%) and testing (40%) subsets to ensure model
generalisability. The RF classifier achieved 80% accuracy on the training set and 59% on the
test set.

A 3-fold cross-validation was also conducted, arriving at a similar 59% accuracy rate.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.70, 0.41 and 0.63 respectively, with precision of 0.77, 0.47 and 0.64 when evaluated with
the test set (consisting of 4394 samples). Lower performance with the Medium cluster
reflects the inherent difficulty of classifying transitional states in natural systems. The
consistent level of accuracy across test and cross-validation demonstrates the generalised
performance of the RF classifier, while not accurate, was deemed sufficiently informative
for the purpose of assessing the influence of reef characteristics.

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its cluster assignment
into Low, Medium, or High. Reef median depth, carrying capacity and mean time series DHW
level have the highest feature importance scores, accounting for 29%, 15% and 15% predictive
importance respectively (@fig-ts-cluster-pdp; Panel A). Reef weighted incoming connectivity,
outgoing connectivity strength, source-to-sink connectivity ratio and larval retention
probability account for \~12%, 11%, 10% and 8% respectively.

@@

Analyses for source-to-sink ratio and incoming and outgoing connectivity metrics suggest
that reefs with greater outgoing connectivity and lower incoming connectivity levels have
higher probability of being clustered into the Low cluster category compared to other reefs.
Additionally, two-way analyses revealed little interaction between depth and incoming and
outgoing connectivity metrics for reef clustering probabilities
(@fig-ts-cluster-pdp-twoway-incoming; @sfig-ts-cluster-pdp-twoway-outgoing). Reefs with a
median depth around 15m and high incoming connectivity may benefit slightly, with higher
probability of High cover cluster assignment compared to 15m - depth reefs with lower
connectivity.

@@

Current reef restoration and intervention efforts on the GBR, such as deployment of coral
seeding, coral fragment devices and Crown of Thorns Starfish control, target reefs with
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, intervention depth is likely constrained by
snorkelling and diving deployment and monitoring logistical accessibility, visual
identification of sites from the surface, in combination with tourism return value. Our
results suggest that of the assessed reefs occurring within the 1-10m depth range (1551
reefs), 59% (920) of reefs are consistently clustered into the Low and Medium cover clusters
across the GCMs (at the GBRMPA management region scale), with many unable to maintain
positive carbonate budgets under future conditions. Partial dependence analysis revealed a
range of depths from 10-15m where the probability of a reef being clustered into the High
cover cluster increased substantially. Additionally, two-way partial dependence analysis
revealed that high incoming connectivity is important for High cluster assignment for reefs
at depths of \>10m. Considering this 10-15m depth range when assessing possible intervention
sites may reduce the exposure of planted and seeded corals to high thermal stress, while
still maintaining effective logistical access to sites. Managers are also likely to consider
the potential for positive flow on effects due to outgoing connectivity when selecting
intervention sites [@hock2017]. Transient, and difficult to measure, properties such as reef
connectivity are likely to increase uncertainty in intervention planning to a greater degree
than environmental characteristics such as reef depth. Overall, reef intervention planning
is a complex and challenging decision-making process [@anthony2020], and our results suggest
that intervention site depth may be an important factor to additionally consider when
assessing the long-term resilience of sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Our results suggest that incoming connectivity contributes positively to high
cover cluster allocation for reefs around 15m deep. However, our results suggest no direct,
first order, relationship between weighted incoming connectivity and the ability of reefs to
maintain positive carbonate budgets under future conditions. There has been contrasting
evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling
is able to identify reefs that are important for recovery processes, with findings dependent
on the larval connectivity model used. These investigations were performed using prior
generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and
1996-2002 [@bode2018] time periods, without considering the influence of reef depth on
resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a
limited temporal sample of particle flows, our results suggest that larval connectivity
alone may not be important or reef resilience and recovery, but should be considered in
combination with more influential factors such as depth. Comprehensive comparisons and
validation of multiple connectivity model outputs would improve certainty and allow for the
targetting of reefs by combining abiotic factors such as depth with connectivity.

@@

The results presented suggest that when modelled as part of a large and complex ecosystem,
reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in
support of depth as a form of thermal refugia for shallow-water reefs on the GBR. Targetting
reefs with median depths greater than 10m may aid in identifying the most resilient reefs
under climate change. We suggest further analyses that investigate the benefits of
incorporating depth-based resilience into reef intervention decision making, as well as
further research and development in connectivity modelling methods for whole-ecosystem
scales.
